Research Experience

### 2010-03-29 doc/thesis/thesis.tex
need to finish the resampling part of particle filters....

%\usepackage{times}
%\usepackage{tikz}
%\usetikzlibrary{trees}
%\usetikzlibrary{shapes,snakes}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{setspace}
\usepackage{graphicx}

@@

\begin{document}
\section*{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
of neural activity. Despite its rather limited temporal resolution, FMRI is still
the best way of measuring neural activity deep in the brain. 
Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis is straight forward.
Because of these benefits, FMRI continues to be crucial to the study of human 
cognition. Despite the fact that FMRI is so widely used,
the choice of analysis tools is still relatively limited. Every widely
available analysis tool is based on parametric modeling, which requires
prior knowledge of the noise distribution. Indeed, the linear methods
usually applied are known not to be robust to non-gaussian, non-white
noise, which is why so many pre-processing steps are necessary.
These limitations cast a long shadow over any experiment that claims to be
able to reject the null hypothesis, no matter the P-value. While obviously
there is a correlation between the BOLD signal and actual activity, its
possible the current parametric tests are over-estimating the actual amount
of activity. Its also possible that the current methods underestimate
actual activity. In this paper we propose a robust model-based approach to 
activation detection that makes no assumptions about the underlying noise
model.

@@

\subsection*{FMRI}
Before going further, it is necessary to discuss Functional MRI,
and what it measures. All MRI works by exciting nuclear spin away
from a steady state controlled by a large electromagnet. Different 
nuclei respond to RF impulses based on their natural frequency
of oscillation. Therefore contrast in MRI is based on the aggregate
nuclear composition of a given region. Thus, regions with more oxygen
will respond differently from regions with proportionately less 
oxygen. This exact difference is what causes the changes over time
in the MR signal in functional MRI. Unfortunately, it takes a significant
amount of time for nuclear spin to return to equilibrium. Thus instead
of waiting, a type of imaging called Echo Planar Imaging (EPI) is used,
wherein a single excitation is used, and then as quickly as possible
every voxel (3 Dimensional Pixel) in a particular region is read. 
Because "reading" a particular region requires the switching on of a
gradient magnetic field, it takes a decent amount of time to perform
measurements. Problematically, the signal from the original excitation
drops exponentially, thus the signal to noise ratio can be very low
even at modest resolutions. Eventually the signal becomes unreadable,
and a new excitation must be applied. As more measurements are made, and thus
as the spatial resolution increases, the rate at which entire volumes
can be acquired drops. Typically volumes can be acquired every two to
three seconds, at resolutions of about 64 by 64 by 32, but their signal
to noise ratio is often poor. Other imaging modalities such as EEG and
MEG are capable of much faster measurements, but are unable to isolate
measurement locations so precisely as FMRI. As with all biological systems,
noise is much higher than in human designed systems, and yet these 
biological systems are capable of performing amazing feats.

\subsection*{BOLD Physiology}
The physiology that leads to the BOLD model has been well studied over the 
past decade and a half. While some uncertainty remains (for instance in
relation to the BOLD post-stimulus undershoot <papers relating to back and
forth on post-stimulus undershoot>) generally the model fits reality very 
well. As with every piece of tissue in the body, the brain requires oxygen
to extract energy from glucose in the blood. This process of removing oxygen
from red blood cells sets into motion a chain of events that locally 
alters the composition of the blood. Inactive neurons can be thought of
as a slingshot cocked and ready to fire. As soon as a signal moves up
the axon and causes the neuron to fire (and thus changes the state
of a membrane at that location), ions quickly move across the altered
membrane to compensate for a high charge imbalance between the interior
and exterior of the cell. This process, similar to allowing a stretched
rubber band to contract leaves the system at a lower energy state. This cannot
last though, because the neuron needs to be ready to fire again rather quickly.
To return the cell to a ready position the membrane becomes impermeable
to the ions again, and then begins pumping ions back into the cell. This
process takes a large amount of energy, whereas the actual firing takes
very little energy. Thus, after firing, glucose is burned, removing oxygen
from the blood and thus causing a dip in the ratio of oxygenated hemoglobin to 
de-oxygenated hemoglobin. This is the first potentially measurable effect
that MRI can see, although it usually lasts fewer than 2 seconds making it
difficult for FMRI to catch. As a result of the decreased amount of local oxygen,
the capillaries compensate by increasing blood flow to that region. Because,
of the quickly increased blood flow into local capillaries, the blood
volume increases in addition to the local oxygen content. This leads to a Windkessel
effect which further lags the normalization of oxygen content. The effect of
all this is an overshoot in the oxygen content above the initial level. After
the work of recharging is done, there may be a prolonged undershoot lasting
as much as 90 seconds \cite{spatial_poststim}, though the reason for this
is debated \cite{origin_poststim}.

@@

\section*{Current Techniques}
\subsection*{Basic Statistical Parametric Mapping}
The most basic method of analyzing FMRI data is through a basic T-test
between "resting state" and "active state" samples. This is done by 
taking the average and variance of the inactive period, and the 
active period separately then treating them both as gaussian distributions.
If they are in fact Gaussian distributions, then a basic t-test will
give the likelihood that the samples came from the same distribution
(the p-value). Of course, this test is fraught with problems; even if
the drift mentioned earlier has been removed, there is little reason
to believe that the noise is Gaussian, or even stable. Additionally, 
even if the noise were Gaussian, a t-test with a p-value of .05 over
5000 or more samples is on average going to generate $.05*5000$ false
positives. To compensate for this, bonferoni correction, also known as
multiple comparison tests are performed; essentially p-values are 
divided by the number of independent
tests being run. This, however, leads to extremely low p-values, so
low that it would be impossible for any biological system to satisfy. To
compensate, a Gaussian kernel is applied to the image, thus reducing
variance (and thus separating the active and inactive distributions)
as well as decreasing the effective number of voxels. Since t-tests are
now no longer being applied to n <I need to define n> independent voxels,
the factor by which the p-value must be divided by can be decreased.
<Do I need to mathematically define all this?> The derivation and application
of random field theory, and its use can be found in various papers \cite{univ_mult_rft}.

\subsection*{General Linear Model}
The most used form of FMRI analysis is still based on Statistical Parametric
Mapping, but is able to account for several different levels or types
of stimulus. By adding a General Linear Model to the analysis, the
output signal timeseries (what FMRI detects) is regressed over the weighted
sum of the various confound's timeseries. The equation for a general linear
model is then
\begin{equation}
Y(t) = X(t)\beta + \epsilon(t)
\end{equation}
where $Y(t)$ is the smoothed or detrended timeseries of measurements,
$X(t)$ is a row vector of stimuli, $\beta$ is a column vector of weights,
and $\epsilon$ is the error. Thus for every time, the measurement is
assumed to be a weighted sum of the inputs plus some error. The calculation
of $\beta$ then is merely a gradient descent search to minimize the
mean squared error.

@@

\subsection*{Whats wrong with these techniques}
There are a few problems with the techniques mentioned in the 
previous sections. First, they essentially ignore prior knowledge about
the system. Although the most advanced form of the general linear model includes
a "Hemodynamic Response Function," that hemodynamic response function is
static across every region of the brain. It is well known that capillary beds
are not uniform and so blood perfusion cannot possible be static across the
brain. Thus, if extra information were available a-priori, that information
could not be incorporated without modifications to the General Linear Model.
Similarly heart rate could not be added either. It would obviously be advantageous
to have true physiological parameters as entry points for these various other
model parameters. The physiological models for the BOLD signal are quite good
and based on realistic physics. While the exact connection between a stimulus
and the flow inducing signal is not precisely known, model fits are actually
quite good \cite{ModelCompareison}. Regardless, being based on some real 
physiological 
parameter would allow for the establishment of reasonable priors and decrease
model variance without breaking a sweat. Of course, using real parameters 
has the additional
bonus of potentially providing information about physical pathologies. It
is quite possible that physical properties such as decreased compliance of
blood vessels could indicate a neurological condition that is not easily
seen in a T1 or T2 map. In essence, this could make FMRI a much more 
useful clinical tool than it is now. The other problem with linear models
is that they are a linear fit to a nonlinear signal. It is not uncommon
for data to be thrown out in FMRI studies because no significant activation
has been seen. However, if, for whatever reason, the BOLD response
was acting more nonlinear than in other patients it would be completely
possible for SPM to miss that activation.

@@

\section*{Proposed Approach}
\subsection*{Goal}
The ultimate goal of this project is to provide a new set of tools
for analyzing FMRI data. Whereas SPM techniques have been highly 
successful at finding macroscopic regions of activation, linear 
modeling can carry significant bias error due to lack of model
flexibility. While adding parameters can significantly increase
error due to model variance, this effect is mitigated by the fact
that we plan to use a model that is based on first principals. The
purpose of this paper is thus to evaluate the potential of using
a particle filter along with the BOLD model to derive physical 
parameters. In so doing, we hope to be able to show that neuronal
efficacy, $\epsilon$ is a suitable variable for estimating voxel 
activation from a standard FMRI image. We also hope to show that 
estimated posterior distribution of the parameters, derived from
the particle filter, is able to provide an accurate measure of the
confidence interval.

\subsection*{Introduction to Particle Filters}
Particle filters, a type of Sequential Monte Carlo (SMC) methods
are a powerful way of estimating the posterior probability distribution
of a set of parameters give a timeseries of measurements. Unlike Markov 
Chain Monte Carlo estimation, Sequential Monte-Carlo methods are designed
to be used with parameters that vary with time. Unlike variations of the
Kalman filter, particle filters do not make the assumption that noise
is Gaussian. Thus particle filters are often the best solution to bayesian 
tracking for non-linear, non-gaussian systems. 
%todo: might need to reference attempts to model a system of our type with
%unscented kalman filter
The idea of the particle
filter is to start with a wide mixture PDF of possible parameter sets, 
and then, as measurements come in, to weight more heavily parameter sets 
that tend to give good estimations of the measurements. The reliance on
an initial mixture PDF can introduce bias; however, this effect can be
minimized by alterring the initial weights in the mixture pdf. Of course
every gradient descent must choose starting points and it is often quite
easy to establish a reasonable range of parameters, especially when the
model being used has a physical meaning. Suppose a set or stream of measurements
are given, $\{y(t), t = 1, 2, 3, ... T\}$, where $T$ is permitted to go
 to infinity. Then the goal is to find the 
parameters, $\hat{\theta}$, and underlying state time series, $\hat{x}[0:T]$
that minimize the difference between $\hat{y}[0:T]$
and $y[0:T]$. In our case, we will assume that we know the form
of the model, which is based on first principals, and that
there is some true $\theta$ and a true time-series of underlying
state variable, $x[0:T]$ that drives $y[0:T]$. Assuming a model form 
such as we do here reduces model variance, potentially at the cost of increased
bias (or systematic) error. We will assume a basic state space model:

@@

\begin{equation}
\dot{x}(t) = f(x(t), u(t)) + \nu_x
\end{equation}

\begin{equation}
y(t) = g(x(t)) + \nu_y
\end{equation}

@@

The goal of the particle filter is to evolve a probably distribution 
$Pr(\hat{x}(T) | u[0:T], y[0:T])$,
that asymptotically approaches the probability distribution $Pr(x(T) | u[0:T])$.
Considering that $y$ contains measurement noise and noise in $x$ can drive
changes in $y$, it is clear that $Pr(x(t) | u[0:T]$ is not a single true value
but a true posterior. 
To begin with, the particle filter starts with a proposal distribution, and $N_p$
particles need to be drawn from that distribution, $\alpha(x)$:
\begin{equation}
\{[\hat{x}_i(0),w_i] : \hat{x}_i(0) \sim \alpha(x), w_i = \frac{1}{N_p}, i \in \{1, 2, ... , N_p\} \}
\end{equation}

Where $N_p$ is the number of particles or points used to describe the prior 
using a Mixture PDF.

\begin{equation}
Pr(\hat{x}(0) = x(0)) = \sum_{i=1}^{N_p} w_i\delta(x(0) - \hat{x}_i(0) ) dx
\end{equation}

Where $\delta(x-x_0)$ is 1 when $x = x_0$ (the Kronecker delta function).

If a true prior is preferred, then the weights
should all be $1/N_p$, and since $\hat{x}_i$ was drawn from the prior, this will
be an approximation of the prior distribution. If a relatively flat prior is 
preferred, then each particle's weight could be divided by the density, $\alpha(\hat{x}_i)$,
which creates a flat prior with support points in the region of $\alpha(x)$. Either
way, $\alpha(x)$ should be much broader than the true posterior, $Pr(x(0))$, since the
choice of support points is crucial to the convergence of any sampling importance
algorithm. 
For the BOLD signal all the parameters have been studied and have relatively well
known mean and variance, so a prior bias is actually preferred. In our
case we did not initially flatten the distributions, but instead used 
distributions that fit what was known about the parameters.  
For strictly positive parameters (members of $\hat{x}$) we used a gamma distribution,
whereas for parameters that could be negative, we used a Gaussian distribution. Both
distributions utilized standard deviations twice that found in previous studies.

Note that all the probability's implicitly depend on $u[0:T]$, so those terms 
will be left off for simplicity.
Once the probability, $Pr(\hat{x}(T) | \hat{x}[0:T-1], y[0:T-1]$ has been found
(initially this is just Mixture approximating the prior since no measurements are 
available and no previous probabilities are available), its possible to approximate
the probability for short times between before a measurement is available, by shifting
the probability according the progression of the state equations. This is only 
an approximate, since integrating $\nu_x$ should increase uncertainty as
time without a measurement passes.

\begin{equation}
Pr(\hat{x}(T+\Delta t)) \approx 
\sum_{i=1}^{N_p} w_i\delta\left(x - (\hat{x}_i(T) + \int_T^{T+\Delta} \dot{\hat{x}}_i(t) dt) \right) dx
\end{equation}

When a measurement becomes available it is incorporated into the probability.
This process of incorporating new data is called sequential importance sampling,
and eventually causes the probability to converge. The weight is defined
as

\begin{equation}
w_i(T) \propto \frac{Pr(x_i[0:T] | y[0:T])}{q(x_i)}
\end{equation}

where $q$ is called an \emph{importance density}, which means it decides where
the support points for $\hat{x}(T)$ are located. To remove the bias due to the
location of the support points, we divide by $q(\hat{x}_i)$. By dividing by 
the density of the support points (particles), the location of the points are removed
from the probability density. As a result the weight is dependent solely based
on $Pr(x_i[0:T] | y[0:T]$, the probability of the $i^{th}$ particle's measurements
being different from $y[0:T]$ due to noise alone.
An example of an importance density would be drawing a large
number of points from the standard normal, $N(0,1)$ and then weighting each point, $l$ by 
$1/\beta(l), \beta \sim N(0,1)$. Of course if there is a far off peak in
the posterior that $q$ does not allocate support points in, there will essentially
be a quantization error, and that part of the density can't be modeled.

For this work we used $q(x_i) = Pr(x_i(T) | x_i[0:T-1]) = Pr(x_i(T) | x_i(T-1))$,
based on the Markov assumption, and the belief that the state space model is 
at least close to correct.  
%<stopped here>
This has the added benefit that it is not necessary to
re-draw points from a $q$ function at every time step, which is faster. 
this the definition of the weight, so when weights are updated it is necessary
to divide by the proposal distributionto converge
to the true distribution it is necessary to divide by the proposal distribution.

\begin{equation}
  d
\end{equation}

>>>>

%\usepackage{times}
%\usepackage{tikz}
%\usetikzlibrary{trees}
%\usetikzlibrary{shapes,snakes}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{cancel}

@@

\begin{document}
\section{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
of neural activity. Despite its rather limited temporal resolution, FMRI is still
the best way of measuring neural activity deep in the brain. 
Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis is straight forward.
Because of these benefits, FMRI continues to be crucial to the study of human 
cognition. Despite the fact that FMRI is so widely used,
the choice of analysis tools is still relatively limited. Every widely
available analysis tool is based on parametric modeling, which requires
prior knowledge of the noise distribution. Indeed, the linear methods
usually applied are known not to be robust to non-gaussian, non-white
noise, which is why so many pre-processing steps are necessary.
These limitations cast a long shadow over any experiment that claims to be
able to reject the null hypothesis, no matter the P-value. While obviously
there is a correlation between the BOLD signal and actual activity, its
possible the current parametric tests are over-estimating the actual amount
of activity. Its also possible that the current methods underestimate
actual activity. In this paper we propose a robust model-based approach to 
activation detection that makes no assumptions about the underlying noise
model.

@@

\subsection{FMRI}
Before going further, it is necessary to discuss Functional MRI,
and what it measures. All MRI works by exciting nuclear spin away
from a steady state controlled by a large electromagnet. Different 
nuclei respond to RF impulses based on their natural frequency
of oscillation. Therefore contrast in MRI is based on the aggregate
nuclear composition of a given region. Thus, regions with more oxygen
will respond differently from regions with proportionately less 
oxygen. This exact difference is what causes the changes over time
in the MR signal in functional MRI. Unfortunately, it takes a significant
amount of time for nuclear spin to return to equilibrium. Thus instead
of waiting, a type of imaging called Echo Planar Imaging (EPI) is used,
wherein a single excitation is used, and then as quickly as possible
every voxel (3 Dimensional Pixel) in a particular region is read. 
Because "reading" a particular region requires the switching on of a
gradient magnetic field, it takes a decent amount of time to perform
measurements. Problematically, the signal from the original excitation
drops exponentially, thus the signal to noise ratio can be very low
even at modest resolutions. Eventually the signal becomes unreadable,
and a new excitation must be applied. As more measurements are made, and thus
as the spatial resolution increases, the rate at which entire volumes
can be acquired drops. Typically volumes can be acquired every two to
three seconds, at resolutions of about 64 by 64 by 32, but their signal
to noise ratio is often poor. Other imaging modalities such as EEG and
MEG are capable of much faster measurements, but are unable to isolate
measurement locations so precisely as FMRI. As with all biological systems,
noise is much higher than in human designed systems, and yet these 
biological systems are capable of performing amazing feats.

\subsection{BOLD Physiology}
The physiology that leads to the BOLD model has been well studied over the 
past decade and a half. While some uncertainty remains (for instance in
relation to the BOLD post-stimulus undershoot <papers relating to back and
forth on post-stimulus undershoot>) generally the model fits reality very 
well. As with every piece of tissue in the body, the brain requires oxygen
to extract energy from glucose in the blood. This process of removing oxygen
from red blood cells sets into motion a chain of events that locally 
alters the composition of the blood. Inactive neurons can be thought of
as a slingshot cocked and ready to fire. As soon as a signal moves up
the axon and causes the neuron to fire (and thus changes the state
of a membrane at that location), ions quickly move across the altered
membrane to compensate for a high charge imbalance between the interior
and exterior of the cell. This process, similar to allowing a stretched
rubber band to contract leaves the system at a lower energy state. This cannot
last though, because the neuron needs to be ready to fire again rather quickly.
To return the cell to a ready position the membrane becomes impermeable
to the ions again, and then begins pumping ions back into the cell. This
process takes a large amount of energy, whereas the actual firing takes
very little energy. Thus, after firing, glucose is burned, removing oxygen
from the blood and thus causing a dip in the ratio of oxygenated hemoglobin to 
de-oxygenated hemoglobin. This is the first potentially measurable effect
that MRI can see, although it usually lasts fewer than 2 seconds making it
difficult for FMRI to catch. As a result of the decreased amount of local oxygen,
the capillaries compensate by increasing blood flow to that region. Because,
of the quickly increased blood flow into local capillaries, the blood
volume increases in addition to the local oxygen content. This leads to a Windkessel
effect which further lags the normalization of oxygen content. The effect of
all this is an overshoot in the oxygen content above the initial level. After
the work of recharging is done, there may be a prolonged undershoot lasting
as much as 90 seconds \cite{spatial_poststim}, though the reason for this
is debated \cite{origin_poststim}.

@@

\section{Current Techniques}
\subsection{Basic Statistical Parametric Mapping}
The most basic method of analyzing FMRI data is through a basic T-test
between "resting state" and "active state" samples. This is done by 
taking the average and variance of the inactive period, and the 
active period separately then treating them both as gaussian distributions.
If they are in fact Gaussian distributions, then a basic t-test will
give the likelihood that the samples came from the same distribution
(the p-value). Of course, this test is fraught with problems; even if
the drift mentioned earlier has been removed, there is little reason
to believe that the noise is Gaussian, or even stable. Additionally, 
even if the noise were Gaussian, a t-test with a p-value of .05 over
5000 or more samples is on average going to generate $.05*5000$ false
positives. To compensate for this, bonferoni correction, also known as
multiple comparison tests are performed; essentially p-values are 
divided by the number of independent
tests being run. This, however, leads to extremely low p-values, so
low that it would be impossible for any biological system to satisfy. To
compensate, a Gaussian kernel is applied to the image, thus reducing
variance (and thus separating the active and inactive distributions)
as well as decreasing the effective number of voxels. Since t-tests are
now no longer being applied to n <I need to define n> independent voxels,
the factor by which the p-value must be divided by can be decreased.
<Do I need to mathematically define all this?> The derivation and application
of random field theory, and its use can be found in various papers \cite{univ_mult_rft}.

\subsection{General Linear Model}
The most used form of FMRI analysis is still based on Statistical Parametric
Mapping, but is able to account for several different levels or types
of stimulus. By adding a General Linear Model to the analysis, the
output signal timeseries (what FMRI detects) is regressed over the weighted
sum of the various confound's timeseries. The equation for a general linear
model is then
\begin{equation}
Y(t) = X(t)\beta + \epsilon(t)
\end{equation}
where $Y(t)$ is the smoothed or detrended timeseries of measurements,
$X(t)$ is a row vector of stimuli, $\beta$ is a column vector of weights,
and $\epsilon$ is the error. Thus for every time, the measurement is
assumed to be a weighted sum of the inputs plus some error. The calculation
of $\beta$ then is merely a gradient descent search to minimize the
mean squared error.

@@

\subsection{Whats wrong with these techniques}
There are a few problems with the techniques mentioned in the 
previous sections. First, they essentially ignore prior knowledge about
the system. Although the most advanced form of the general linear model includes
a "Hemodynamic Response Function," that hemodynamic response function is
static across every region of the brain. It is well known that capillary beds
are not uniform and so blood perfusion cannot possible be static across the
brain. Thus, if extra information were available a-priori, that information
could not be incorporated without modifications to the General Linear Model.
Similarly heart rate could not be added either. It would obviously be advantageous
to have true physiological parameters as entry points for these various other
model parameters. The physiological models for the BOLD signal are quite good
and based on realistic physics. While the exact connection between a stimulus
and the flow inducing signal is not precisely known, model fits are actually
quite good \cite{ModelCompareison}. Regardless, being based on some real 
physiological 
parameter would allow for the establishment of reasonable priors and decrease
model variance without breaking a sweat. Of course, using real parameters 
has the additional
bonus of potentially providing information about physical pathologies. It
is quite possible that physical properties such as decreased compliance of
blood vessels could indicate a neurological condition that is not easily
seen in a T1 or T2 map. In essence, this could make FMRI a much more 
useful clinical tool than it is now. The other problem with linear models
is that they are a linear fit to a nonlinear signal. It is not uncommon
for data to be thrown out in FMRI studies because no significant activation
has been seen. However, if, for whatever reason, the BOLD response
was acting more nonlinear than in other patients it would be completely
possible for SPM to miss that activation.

@@

\section{Proposed Approach}
\subsection{Goal}
The ultimate goal of this project is to provide a new set of tools
for analyzing FMRI data. Whereas SPM techniques have been highly 
successful at finding macroscopic regions of activation, linear 
modeling can carry significant bias error due to lack of model
flexibility. While adding parameters can significantly increase
error due to model variance, this effect is mitigated by the fact
that we plan to use a model that is based on first principals. The
purpose of this paper is thus to evaluate the potential of using
a particle filter along with the BOLD model to derive physical 
parameters. In so doing, we hope to be able to show that neuronal
efficacy, $\epsilon$ is a suitable variable for estimating voxel 
activation from a standard FMRI image. We also hope to show that 
estimated posterior distribution of the parameters, derived from
the particle filter, is able to provide an accurate measure of the
confidence interval.

\subsection{Introduction to Particle Filters}
Particle filters, a type of Sequential Monte Carlo (SMC) methods
are a powerful way of estimating the posterior probability distribution
of a set of parameters give a timeseries of measurements. Unlike Markov 
Chain Monte Carlo estimation, Sequential Monte-Carlo methods are designed
to be used with parameters that vary with time. Unlike variations of the
Kalman filter, particle filters do not make the assumption that noise
is Gaussian. Thus particle filters are often the best solution to bayesian 
tracking for non-linear, non-gaussian systems. 
%todo: might need to reference attempts to model a system of our type with
%unscented kalman filter
\subsubsection{Model}
The idea of the particle
filter is to start with a wide mixture PDF of possible parameter sets, 
and then, as measurements come in, to weight more heavily parameter sets 
that tend to give good estimations of the measurements. The reliance on
an initial mixture PDF can introduce bias; however, this effect can be
minimized by alterring the initial weights in the mixture pdf. Of course
every gradient descent must choose starting points and it is often quite
easy to establish a reasonable range of parameters, especially when the
model being used has a physical meaning. Suppose a set or stream of measurements
are given, $\{y(t), t = 1, 2, 3, ... T\}$, where $T$ is permitted to go
 to infinity. Then the goal is to find the 
parameters, $\hat{\theta}$, and underlying state time series, $\hat{x}[0:T]$
that minimize the difference between $\hat{y}[0:T]$
and $y[0:T]$. In our case, we will assume that we know the form
of the model, which is based on first principals, and that
there is some true $\theta$ and a true time-series of underlying
state variable, $x[0:T]$ that drives $y[0:T]$. Assuming a model form 
such as we do here reduces model variance, potentially at the cost of increased
bias (or systematic) error. We will assume a basic state space model:

@@

\begin{equation}
\label{stateass}
\dot{x}(t) = f(x(t), u(t)) + \nu_x
\end{equation}

\begin{equation}
\label{measass}
y(t) = g(x(t)) + \nu_y
\end{equation}

@@

\subsubsection{Prior}

The goal of the particle filter is to evolve a probably distribution 
$Pr(\hat{x}(T) | u[0:T], y[0:T])$,
that asymptotically approaches the probability distribution $Pr(x(T) | u[0:T])$.
Considering that $y$ contains measurement noise and noise in $x$ can drive
changes in $y$, it is clear that $Pr(x(t) | u[0:T])$ is not a single true value
but a true posterior. 
To begin with, the particle filter starts with a proposal distribution, and $N_p$
particles need to be drawn from that distribution, $\alpha(x)$:
\begin{equation}
\{\hat{Pr}x_i(0),w_i] : x_i(0) \sim \alpha(x), w_i = \frac{1}{N_p}, i \in \{1, 2, ... , N_p\} \}
\end{equation}

Where $N_p$ is the number of particles or points used to describe the prior 
using a Mixture PDF. 
\begin{equation}
\hat{Pr}(x(0) = \hat{x}) = \sum_{i=1}^{N_p} w_i\delta(\hat{x} - x_i(0) ) dx
\end{equation}
Where $\delta(x-x_0)$ is 1 if and only if $x = x_0$ (the Kronecker delta function).

If a true prior is preferred, then the weights
should all be $1/N_p$, and since $x_i$ was drawn from the prior, this will
be an approximation of the prior distribution. If a relatively flat prior is 
preferred, then each particle's weight could be divided by the density, $\alpha(x_i)$,
which creates a flat prior with support points in the region of $\alpha(x)$. Either
way, $\alpha(x)$ should be much broader than the true posterior, $Pr(x(0))$, since the
choice of support points is crucial to the convergence of any sampling importance
algorithm. 
For the BOLD signal all the parameters have been studied and have relatively well
known mean and variance, so a prior could be very helpful. We ran simulations for
both normalized and un-normalized priors, although we believe in cases such as this,
where a good prior exists, it should be used. 
For strictly positive parameters (members of $x$) we used a gamma distribution,
whereas for parameters that could be negative, we used a Gaussian distribution. In
both cases standard deviations twice that found in previous studies were used.

Note that all the probabilities implicitly depend on $u[0:T]$, so those terms 
will be left off for simplicity.
Once the probability, $\hat{Pr}(x(T) | x[0:T-1], y[0:T-1])$ has been found
(initially this is just Mixture approximating the prior since no measurements are 
available and no previous probabilities are available), its possible to approximate
the probability for short times between times when measurement is available, by shifting
the probability according the progression of the state equations. This is only 
an approximate, since integrating $\nu_x$ should increase uncertainty as
time without a measurement passes.

\begin{equation}
\hat{Pr}(x(T+\Delta t)) \approx 
\sum_{i=1}^{N_p} w_i\delta\left(x - (x_i(T) + \int_T^{T+\Delta} \dot{x}_i(t) dt) \right)
\end{equation}

\subsubsection{Weighting}
When a measurement becomes available it is incorporated into the probability.
This process of incorporating new data is called sequential importance sampling,
and eventually causes the probability to converge. The weight is defined
as

\begin{equation}
\label{weightfunc}
w_i(T) \propto \frac{\hat{Pr}(x_i[0:T] | y[0:T])}{q(x_i[0:T] | y[0:T])}
\end{equation}

where $q$ is called an \emph{importance density}, meaning it decides where
the support points for $x(T)$ are located. To remove the bias due to the
location of the support points, we divide by $q(x_i[0:T] | y[0:T])$. By dividing by 
the posterior density of the support points (particles), the effect of the particle 
distribution may be removed from the posterior density. As a result the weight
is dependent solely based
on $\hat{Pr}(x_i[0:T] | y[0:T])$, the probability of the $i^{th}$ particle's measurements
being different from $y[0:T]$ due to noise alone.
An example of an importance density would be drawing a large
number of points from the standard normal, $N(0,1)$ and then weighting each point, $l$ by 
$1/\beta(l), \beta \sim N(0,1)$. Of course if there is a far off peak in
the posterior that $q$ does not allocate support points in, there will 
be a quantization error, and that part of the density can't be modeled. This is why
it is absolutely necessary that $q$ covers $\hat{Pr}(x_i[0:T] | y[0:T])$.

$q(x_i[0:T] | y[0:T])$ may be simplified by assuming that $y(T)$ doesn't contain 
any information about $x(T-1)$, which is more practical since knowledge of future
measurements is impractical.

\begin{eqnarray}
q(x[0:T] | y[0:T]) & = & q(x(T) | x[0:T-1], y[0:T])q(x[0:T-1] | y[0:T]) \nonumber \\
& = & q(x(T) | x[0:T-1], y[0:T])q(x[0:T-1] | y[0:T-1]) \nonumber \\
& = & q(x(T) | x(T-1), y[0:T])q(x[0:T-1] | y[0:T-1])
\end{eqnarray}

In this paper we will use 
$q(x_i(T) | x_i(T-1), y[0:T]) =  \hat{Pr}(x_i(T) | x_i(T-1))$,
based on the Markov assumption, and the belief that the state space model is 
able to approximate the true state. This means that prior to re-weighting 
particles, the particles will be distributed the same as the previous time but
moved forward according to the integration of $f(x(t), u(t))$.

In addition to $q(x_i(T) | x_i[0:T-1], y[0:T])$, the weight is also based on $Pr(x_i[0:K] | y[0:K])$,
which may be broken up as follows.
\begin{eqnarray}
\hat{Pr}(x[0:T] | y[0:T]) & = & \frac{\hat{Pr}(y[0:T], x[0:T])}{\hat{Pr}(y[0:T])} \nonumber \\
 & = & \frac{\hat{Pr}(y(T), x[0:T] | y[0:T-1]) \cancel{\hat{Pr}(y[0:T-1])}}{\hat{Pr}(y(T) | y[0:T-1]) \cancel{\hat{Pr}(y[0:T-1])}} \nonumber \\
 & = & \frac{\hat{Pr}(y(T)| x[0:T], y[0:T-1]) \hat{Pr}(x[0:T] | y[0:T-1])}{\hat{Pr}(y(T) | y[0:T-1]) } \nonumber \\
 & = & \frac{\hat{Pr}(y(T)| x[0:T], y[0:T-1]) \hat{Pr}(x(T) | x[0:T-1], y[0:T-1]) \hat{Pr}(x[0:T-1] | y[0:T-1])}{\hat{Pr}(y(T) | y[0:T-1])} \nonumber \\
\end{eqnarray}

Using the assumption that $y(t)$ is fully constrained by $x(t)$ \eqref{measass},
and that $x(t)$ is fully constrained by $x(t-1)$ \eqref{stateass}, we are able to
make the reasonably good assumptions that:
\begin{equation}
\hat{Pr}(y(T) | x[0:T], y[0:T-1]) = \hat{Pr}(y(T) | x(T))
\end{equation}

\begin{equation}
\hat{Pr}(x(T) | x[0:T], y[0:T-1]) = \hat{Pr}(x(T) | x(T-1))
\end{equation}

Additionally, for the particle filter $y(T)$ and $y[0:T-1]$ are 
given, and therefore constant across all particles. Thus $\hat{Pr}(x[0:T] | y[0:T])$
may be simplified to:
\begin{eqnarray}
\hat{Pr}(x[0:T] | y[0:T]) & = & \frac{\hat{Pr}(y(T)| x[0:T], y[0:T-1]) \hat{Pr}(x(T) | x[0:T-1], y[0:T-1]) 
            \hat{Pr}(x[0:T-1] | y[0:T-1])}{\hat{Pr}(y(T) | y[0:T-1])} \nonumber \\
& = & \frac{\hat{Pr}(y(T)| x(T)) \hat{Pr}(x(T) | x(T-1)) \hat{Pr}(x[0:T-1] | y[0:T-1])}{\hat{Pr}(y(T) | y[0:T-1])} \nonumber \\
& \propto & \hat{Pr}(y(T)| x(T)) \hat{Pr}(x(T) | x(T-1)) \hat{Pr}(x[0:T-1] | y[0:T-1])
\end{eqnarray}

Plugging these simplifications into \eqref{weightfunc} leads to:
\begin{eqnarray}
\label{weightevolve}
w_i(T) & \propto & \frac{\hat{Pr}(y(T)| x(T)) \cancel{\hat{Pr}(x(T) | x(T-1))} \hat{Pr}(x[0:T-1] | y[0:T-1])}
                         {\cancel{\hat{Pr}(x_i(T) | x_i(T-1))}q(x[0:T-1] | y[0:T-1])} \nonumber \\
& \propto & w_i(T-1)\hat{Pr}(y(T)| x(T)) 
\end{eqnarray}

Thus, by making the following relatively weak assumptions, evolving a posterior
density  is easy and requires almost no knowledge of noise distribution.
\begin{enumerate}
\item $f(t, x(t), u(t)) = f(x(t), u(t))$ and $g(t, x(t), u(t)) = g(x(t))$ provide 
a sufficiently flexible model to encapsulate the true time series.
\item $E[\nu_x] = 0$ and $E[\nu_y] = 0$, and $\nu_x$ and $\nu_y$ are stationary
\item The PDF $q(x_i(0))$ (the prior) fully covers $Pr(x_i(0))$
\item Markov Assumption: $Pr(x(T) | x[0:T]) = Pr(x(T) | x(T-1))$
\item $q(x[0:T-1] | y[0:T]) = q(x[0:T-1] | y[0:T-1])$
\end{enumerate}

\subsubsection{Basic Particle Filter Algorithm}
From the definition of $w_i$, the algorithm sequential importance sampling (SIS) 
is relatively simple.
